On the use of relative brain size

### 2022-07-11 manuscript.Rmd
Updating main text. Only minor changes.

---
title: "On the use of relative brain size"
author: "Simeon Q. Smeele $^{1,2,3}$"
output:
  pdf_document: default
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{setspace}
- \doublespacing
editor_options:
  chunk_output_type: console
bibliography: bibliography.bib
---

@@

More recently another statistical caveat has been highlighted (for an overview see @freckleton2002misuse). The use of residuals leads to biased estimates if correlation exists between the predictor variables, as is often the case in the comparative literature. The proposed solution is to include body size in a multiple regression, which is very similar to the use of residual brain size, if - and only if - relative brain size is the response variable. This approach takes care of both the phylogenetic signal (since the model can include the phylogenetic variance-covariance matrix) and correct estimation of uncertainty. However, one major issue has been overlooked: if relative brain size is a predictor variable, including body size as a second predictor leads to incorrect inference. This becomes clear when looking at the causal structure of such a case. The best way to decide which variables to include in an analysis is the use of causal graphs [@glymour2016causal, @wright1934method]. @laubach2021biologist presented a simple guide to draw Directed Acyclic Graphs (DAG), in which all variables are connected with arrows that indicate the direction of causality.

To point out the difference between the case with relative brain size as response vs predictor variable, I will present the underlying assumed causal structure presented in Figure 1. To avoid discussion about directionality I have simply named the third variable *z*. In the first case both body size and *z* cause brain size (see Figure 1a). To decide how to best estimate the direct effect of *z* on brain size, one needs to make sure that all *back-door paths* are closed. In other words, all arrows that points towards *z* need to be considered. But since there are no such arrows in this case, there is no need to include additional variables. However, body size explains most of the variation in brain size. Therefore it is still a good idea to include this variable to get a more precise estimate of the effect of *z* on brain size. Or in other words on relative brain size, since allometry is now accounted for.

@@

I analysed the resulting data with a frequentist linear model and with the Bayesian equivalent. Then I plotted the estimated coefficient of body size and *z* to show how well parameters were retrieved. Additionally the bias (difference between estimate and simulated value) and coverage (proportion of times the true value was within 2SE distance from the estimate) are reported.

@@

The model includes a regression for each variable. Body size is not a function of any variable. Brain size is a function of body size. *z* is a function of body size and relative brain size (where relative brain size is the difference between the actual and predicted brain size). Relative brain size in this model is very similar to residual brain size, but since it is computed at each iteration information flows in both directions and measurement error is correctly estimated. The last three lines of the model are the priors for all parameters. Note that the priors for the slopes ($\beta$ and $\gamma$) are set to normal(0, 1), which regularises them slightly and are centred no effect of the predictors. For empirical studies theory might provide more informative priors, which would further increase the accuracy of the model.

@@

# Calculate average correlation between body size and absolute brain size
correlation = sapply(out, function(x) cor(x$body_size, x$brain_size))
mean_correlation = mean(correlation)
```

@@

```{r table, echo=FALSE}
table_combined = rbind(table_response, table_predictor)
knitr::kable(table_combined, format = 'markdown', caption = 'Bias and coverage of all models. Mean and standard deviation bias is calculated from the distance between the point estimate and simulated value. Coverage is the proportion of times that the simulated value was withing the point estimate +- 2SE.')
```

@@

Several recent studies have claimed to study the effect of relative brain size by including both absolute brain size and body size as predictors [@street2017coevolution; @gonzalez2010large; @isler2009why, @maklakov2011brains, @samia2015brain]. These studies also reported the direct effect of body size and sometimes drew conclusions based on the sign of this effect. Since all these studies used some version of a linear model (be it phylogenetic and/or Bayesian), they actually tested the effect of absolute brain size, since including body size only accounts for allometry in the response variable. The simulations in this paper showed that including body size as additional predictor to control for allometric scaling of brain size works well if relative brain size is the response variable, but not if relative brain size is a predictor variable. Perhaps counter-intuitively, the effect of brain size was still estimated correctly by all models. It was the body size effect that was biased in the linear models. Simulations with a strong body size effect still found a positive coefficient for this effect, but were biased towards lower values (see Figure 12). In simulations with a strong brain size effect the body size coefficients were even negative (see Figure 13).

@@

The use of an artificial variable *z* might be confusing, so it might be informative to give examples for both cases and explain the evolutionary causal structure. The social brain hypothesis is possible the most hotly debated version of case I, with relative brain size as response variable. Here brain size is a function of body size and some measure of sociality (*z*). The idea is that social species experience evolutionary pressure to increase brain size to deal with complex social situations. First brought up by @jolly1966lemur, this hypotheses has since received much attention both with residual brain size and absolute brain size as response variable (for a review see @acedo2016critical). The second case, where relative brain size is a predictor variable, has received less attention, but several studies have attempted to show that relative brain size can cause longevity, with relatively larger brained species living longer [@street2017coevolution, @yu2018large, @decasien2018encephalization, @jimenez2020long]. The understanding is that cognitive ability causes both increased brain size and the ability to survive longer. This might be unintuitive, since mechanistically cognition is caused by the brain, but in an evolutionary context it makes sense, since species that require more cognitive flexibility experience evolutionary pressure to increase brain size.

For this paper I drew all variables from normal distributions with zero mean and standard deviation one. I furthermore simulated *z* as a function of relative brain size, rather than simulating both *z* and brain size as a function of the unobserved cognitive ability (as depicted in Figure 1b). I chose to do this to illustrate that even under simple conditions a multiple linear regression cannot estimate the effect of body size correctly. In empirical studies the exact causal structure will be different, but the general observation that one cannot control for allometry in a multiple linear regression when brain size is not the response variable still stands. A regular confounder check as described by @laubach2021biologist should inform the design of the model. One example of relative brain size as predictor variable is the study of life history and innovativeness. @sol2016life suggested that relative brain size predicts both the innovation propensity and maximum life span of birds. They recognised the critique on using residuals, but still used this approach, because they wanted to remove the allometric effects from brain size and not from innovation propensity. This was a valid argument since the original critique assumed a case where $x_1$ and $x_2$ are both causing $y$ [@freckleton2002misuse]. When $x_1$ and $x_2$ are correlated, the use of a residual of $y \sim x_1$ creates a bias. In the case of life history and innovativeness the causal structure is different and this particular problem does not arise. However, in an empirical study measurement error, missing data and phylogenetic covariance still need to be accounted for. With the use of a structural equation model one obtains direct estimates of relative brain size, which includes uncertainty stemming from measurement uncertainty on body size and on brain size. This can also be plotted directly, allowing researchers to visualise how certain the correlation of relative brain size with the third variable is.

@@

# Supplemental methods

## Sample size

@@

## Priors

@@

## Effect magnitude

@@

# Supplemental results

>>>>

---
title: "On the use of relative brain size"
author: "Simeon Q. Smeele $^{1,2,3}$"
output:
  word_document: default
  pdf_document: default
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{setspace}
- \doublespacing
editor_options:
  chunk_output_type: console
bibliography: bibliography.bib
---

@@

More recently another statistical caveat has been highlighted (for an overview see @freckleton2002misuse). The use of residuals leads to biased estimates if correlation exists between the predictor variables, as is often the case in the comparative literature. The proposed solution is to include body size in a multiple regression, which is very similar to the use of residual brain size, if - and only if - relative brain size is the response variable. This approach takes care of both the phylogenetic signal (since the model can include the phylogenetic variance-covariance matrix) and correct estimation of uncertainty. However, one major issue has been overlooked: if relative brain size is a predictor variable, including body size as a second predictor leads to incorrect inference. This becomes clear when looking at the causal structure of such a case. The best way to decide which variables to include in an analysis is the use of causal graphs [@glymour2016causal; @wright1934method]. @laubach2021biologist presented a simple guide to draw Directed Acyclic Graphs (DAG), in which all variables are connected with arrows that indicate the direction of causality.

To point out the difference between the case with relative brain size as response vs predictor variable, I will present the underlying assumed causal structure visualised in Figure 1. To avoid discussion about directionality I have simply named the third variable *z*. In the first case both body size and *z* cause brain size (see Figure 1a). To decide how to best estimate the direct effect of *z* on brain size, one needs to make sure that all *back-door paths* are closed. In other words, all arrows that points towards *z* need to be considered. But since there are no such arrows in this case, there is no need to include additional variables. However, body size explains most of the variation in brain size. Therefore it is still a good idea to include this variable to get a more precise estimate of the effect of *z* on brain size. Or in other words on relative brain size, since allometry is now accounted for.

@@

I analysed the resulting data with a frequentist linear model and with the Bayesian equivalent. Then I plotted the estimated coefficient of body size and *z* to show how well parameters were retrieved. Additionally I reported the bias (difference between estimate and simulated value) and coverage (proportion of times the true value was within 2SE distance from the estimate).

@@

The model includes a regression for each variable. Body size is not a function of any variable. Brain size is a function of body size. *z* is a function of body size and relative brain size (where relative brain size is the difference between the actual and predicted brain size). Relative brain size in this model is very similar to residual brain size, but since it is computed at each iteration information flows in both directions and measurement error is correctly estimated. The last three lines of the model are the priors for all parameters. Note that the priors for the slopes ($\beta$ and $\gamma$) are set to normal(0, 1), which regularises them slightly and center them around no effect of the predictors. For empirical studies theory might provide more informative priors, which would further increase the accuracy of the model.

@@

# Calculate average correlation between body size and absolute brain size
correlation = sapply(out, function(x) cor(x$body_size, x$brain_size))
mean_correlation = round(mean(correlation), 2)
```

@@

```{r table, echo=FALSE}
table_combined = rbind(table_response, table_predictor)
knitr::kable(table_combined, format = 'markdown', caption = 'Bias and coverage of all models. The mean and standard deviation of the bias is calculated from the distance between the point estimate and simulated value. Coverage is the proportion of times that the simulated value was withing the point estimate +- 2SE.')
```

@@

Several recent studies have claimed to study the effect of relative brain size by including both absolute brain size and body size as predictors [@street2017coevolution; @gonzalez2010large; @isler2009why; @maklakov2011brains; @samia2015brain]. These studies also reported the direct effect of body size and sometimes drew conclusions based on the sign of this effect. Since all these studies used some version of a linear model (be it phylogenetic and/or Bayesian), they actually tested the effect of absolute brain size, since including body size only accounts for allometry in the response variable. The simulations in this paper showed that including body size as additional predictor to control for allometric scaling of brain size works well if relative brain size is the response variable, but not if relative brain size is a predictor variable. Perhaps counter-intuitively, the effect of brain size was still estimated correctly by all models. It was the body size effect that was biased in the linear models. Simulations with a strong body size effect still found a positive coefficient for this effect, but were biased towards lower values (see Figure 12). In simulations with a strong brain size effect the body size coefficients were even negative (see Figure 13).

@@

The use of an artificial variable *z* might be confusing, so it might be informative to give examples for both cases and explain the evolutionary causal structure. The social brain hypothesis is possibly the most hotly debated version of case I, with relative brain size as response variable. Here brain size is a function of body size and some measure of sociality (*z*). The idea is that social species experience evolutionary pressure to increase brain size to deal with complex social situations. First brought up by @jolly1966lemur, this hypotheses has since received much attention both with residual brain size and absolute brain size as response variable (for a review see @acedo2016critical). The second case, where relative brain size is a predictor variable, has received less attention, but several studies have attempted to show that relative brain size can cause longevity, with relatively larger brained species living longer [@street2017coevolution; @yu2018large; @decasien2018encephalization; @jimenez2020long]. The understanding is that cognitive ability causes both increased brain size and the ability to survive longer. This might be unintuitive, since mechanistically cognition is caused by the brain, but in an evolutionary context it makes sense, since species that require more cognitive flexibility experience evolutionary pressure to increase brain size.

For this paper I drew all independent variables from normal distributions with zero mean and standard deviation one. I furthermore simulated *z* as a function of relative brain size, rather than simulating both *z* and brain size as a function of the unobserved cognitive ability (as depicted in Figure 1b). I chose to do this to illustrate that even under simple conditions a multiple linear regression cannot estimate the effect of body size correctly. In empirical studies the exact causal structure will be different, but the general observation that one cannot control for allometry in a multiple linear regression when brain size is not the response variable still stands. A regular confounder check as described by @laubach2021biologist should inform the design of the model. One example of relative brain size as predictor variable is the study of life history and innovativeness. @sol2016life suggested that relative brain size predicts both the innovation propensity and maximum life span of birds. They recognised the critique on using residuals, but still used this approach, because they wanted to remove the allometric effects from brain size and not from innovation propensity. This was a valid argument since the original critique assumed a case where $x_1$ and $x_2$ are both causing $y$ [@freckleton2002misuse]. When $x_1$ and $x_2$ are correlated, the use of a residual of $y \sim x_1$ creates a bias. In the case of life history and innovativeness the causal structure is different and this particular problem does not arise. However, in an empirical study measurement error, missing data and phylogenetic covariance still need to be accounted for. With the use of a structural equation model one obtains direct estimates of relative brain size, which includes uncertainty stemming from measurement uncertainty on body size and on brain size. This can also be plotted directly, allowing researchers to visualise how certain the correlation of relative brain size with the third variable is.

@@

## Supplemental methods

### Sample size

@@

### Priors

@@

### Effect magnitude

@@

## Supplemental results
